Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-02 doc/paper-lcr.tex
started re-write of theory

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For a strong lens system in an otherwise
homogeneous Robertson--Walker universe, the excess time delay of an image at angular position
$\bmath{\theta}=(\theta_1,\theta_2)$ with unlensed source position $\bmath{\beta}=(\beta_1,\beta_2)$ is
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distance from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right]. 
\ee
The first term in Eq. \ref{eq:FP} comes from the geometric path difference as a result of
the deflection, and the second term is the gravitational delay
described by the lens potential $\psi(\bmath{\theta})$.

@@

\subsection{Mass-sheet degeneracy}
\label{sec:H0theory:MSD}

A dificulty with using time delay distances to contrain $H_0$ is the so called
Mass-sheet degeneracy \citep[see e.g.][for details]{Falco1985,Schneider2006}.
This is a degeneracy in the strong lens modeling corresponding to the addition of a
mass sheet that contributes a convergence and zero shear which combined with a
re-scaling of the original mass distribution leaves the predicted image positions and
magnifications unchanged. Given a lens
model $\kappa_{\rm lens}(\bmath{\theta})$, an alternative model described by the
transformation 
\be\label{eq:transform}
\widetilde{\kappa}_{\rm lens}(\bmath{\theta}) = \lambda + (1-\lambda)\kappa_{\rm lens}(\bmath{\theta}),
\ee
where $\lambda$ is a constant,would fit the image positions, shapes and relative fluxes equally well. Physically $\lambda$
is the \flag{efffective?} convergence of the mass-sheet and can take any real value. The mass-sheet degeneracy
corresponds to a degeneracy between this external convergence ($\kappax$) and the mass
normalization of the lens galaxy.

Since strong lenses sit in our real Universe, they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. It is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and precisely \citep{suyu2010}.

Despite the invariance of the image positions, shapes and relative fluxes
under the transformation in Eq. \ref{eq:transform}, the relative Fermat potential between the
images is not invariant, but changes according to
\be\label{eq:phitransform}
\Delta \widetilde{\phi_{\rm }}(\bmath{\theta},\widetilde{\bmath{\beta}_{\rm }}) 
= 
(1-\lambda) \Delta \phi_{\rm }
(\bmath{\theta},\bmath{\beta}_{\rm }).
\ee
If the source brightness changes the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. These time delays are inversely proportional to $H_0$ and
proportional to the relative Fermat potential (Equation \ref{eq:Tsimp}). Thus if
there is any external convergence, $\kappax$, that is not included in the lens
modeling, then the value of $H_0$ --inferred assuming $\kappax = 0$ -- will be
$1-\kappax$ less than the true value of $H_0$:
\be 
\label{eq:MassSheet:H0bias}
H_0^{\rm{true}}=(1-\kappax) H_0^{{\kappax = 0}}.
\ee
This degeneracy is present because lensing observations only deliver
relative positions and fluxes. However we can overcome this degeneracy 
if we have knowledge of $\kappax$.

For mass-sheets that are physically
associated with the lens galaxy, the mass sheet affects the stellar dynamics of the
lens galaxy. Stellar dynamics can be used jointly with lensing to break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's mass
\citep[e.g.,][]{GroginNarayan96a, GroginNarayan96b, TonryFranx99, KoopmansTreu02,
 TreuKoopmans02, BarnabeKoopmans07, suyu2010}.

@@

Weak lensing by NFW halos. Accounting for mass along the line of sight.
Approximations to full ray tracing: combining kappas: simple sum, Keeton
scaling, comments. Example ray traced kappa map from Stefan.

>>>>

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from calculated from 
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distances from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is given by
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta}$ is the lens potential.
%The two term in Eq. \ref{eq:FP} comes from the geometric path difference as a result of
%the deflection, and the second term is the gravitational delay
%described by the lens potential $\psi(\bmath{\theta})$.

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes. The mass-sheet degeneracy
\citep[see e.g.][for details]{Falco1985,Schneider2006}. This is a degeneracy in the strong lens
modeling corresponding to the addition of a mass sheet that contributes a convergence and
zero shear which, combined with a re-scaling of the original mass distribution, leaves the
predicted image positions and magnifications unchanged.
%Given a lens
%model $\kappa_{\rm lens}(\bmath{\theta})$, an alternative model described by the
%transformation 
%\be\label{eq:transform}
%\widetilde{\kappa}_{\rm lens}(\bmath{\theta}) = \lambda + (1-\lambda)\kappa_{\rm lens}(\bmath{\theta}),
%\ee
%where $\lambda$ is a constant, would fit the image positions, shapes and relative fluxes equally well. Physic%ally $\lambda$
%is the \flag{effective?} convergence of the mass-sheet and can take any real value. The mass-sheet degeneracy
%corresponds to a degeneracy between this external convergence ($\kappax$) and the mass
%normalization of the lens galaxy.
If the source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and
precisely \citep{suyu2010} from observed time delays.

\be\label{eq:phitransform}
\Delta \widetilde{\phi_{\rm }}(\bmath{\theta},\widetilde{\bmath{\beta}_{\rm }}) 
= 
(1-\lambda) \Delta \phi_{\rm }
(\bmath{\theta},\bmath{\beta}_{\rm }).
\ee
 Thus if
there is any external convergence, $\kappax$, that is not included in the lens
modeling, then the value of $H_0$ --inferred assuming $\kappax = 0$ -- will be
$1-\kappax$ less than the true value of $H_0$:
\be 
\label{eq:MassSheet:H0bias}
H_0^{\rm{true}}=(1-\kappax) H_0^{{\kappax = 0}}.
\ee
This degeneracy is present because lensing observations only deliver
relative positions and fluxes. However we can overcome this degeneracy 
if we have knowledge of $\kappax$.

@@

For mass-sheets that are physically
associated with the lens galaxy, the mass sheet affects the stellar dynamics of the
lens galaxy, so dynamical observations can break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's mass
\citep[e.g.,][]{GroginNarayan96a, GroginNarayan96b, TonryFranx99, KoopmansTreu02,
 TreuKoopmans02, BarnabeKoopmans07, suyu2010}.

@@

For real lines of sight, it is nearly impossible to know the true $\kappax$ for a
line of sight, and without a knowledge of $\kappax^{\mathrm{true}}$ it is hard to
quantify if there is a bias induced by using different $\kappax$ estimators.
A source with known unlensed brightness e.g. a strongly lensed type Ia
Supernova \flag{cite some SNe papers} or a source of known unlensed size \flag{ideas?}
would break the mass-sheet degeneracy and provide a value for $\kappax$, but
\flag{no?} such systems are currently known.

Cosmological dark matter simulations can be used to generate mock lines of sight...
\flag{Stefan's section on how halos are extracted from the MS n-body particles}

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW halos \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s} = r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
 and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are more concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. $\rho_0$ is calculated 
by the critical density and concentration parameter:
\be
$\rho_0$ = \rho_{\mathrm{crit}}(z\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee

From the NFW density profile, \citet{wrightbrainerd1999} derive the following quantities
which are necessary for our line of sight reconstruction:

The mass enclosed within radius is given by:

The projected mass density is given by:

The shear is given by:

Weak lensing by NFW halos. Accounting for mass along the line of sight.
Approximations to full ray tracing:
